Prenatal conditions do not affect brain physiology and learning in a lizard

### 2025-04-20 ms.qmd
updates

---
title: "Prenatal conditions do not affect brain physiology and learning in a lizard"
docx: 
bibliography: "./bib/refs.bib"
csl: "./bib/biology-letters.csl"
reference-doc: "./bib/tmpl.docx"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Figure    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Figure   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

@@

We fitted a multivariate structural equation model (SEM) using _brms_ to explore direct and indirect links between mitochondrial function and learning. The model was structured based on theoretical expectations shown in @fig-sem. Any missing data was imputed during model fitting using data augmentation [@noble2021planned], but this was largely restricted to DNA damage and lipid peroxidation. To obtain a measure of learning performance we extracted the posterior distribution of individual learning slopes (i.e., changes in error over time) by including a random trial slope for each lizard. Learning slopes and the rest of variables, were standardized as before (see above). We assumed a Gaussian error distribution for all the variables in this model. Factors found to be non-significant in the univariate models were excluded. Clutch identity remained a random factor in our SEM model. Direct, indirect, and total effects were derived from posterior estimates [@kline2005principles] (see table S4).

@@

We found that the lizards improved their choices over time, making less mistakes as the trials progressed (see @fig-learning and table S1). However, the learning rate was not influenced by CORT, temperature, or their interaction (see contrasts in table S2).

Mitochondrial density, metabolic capacity, ROS, DNA damage, and lipid peroxidation were also not affected by prenatal conditions (see @fig-results_mit and table S2). We did not find effects of sex and age on mitochondrial physiology (see table S3 and tables S5-S9).

@@

Our SEM showed that mitochondrial physiology was unrelated to learning abilities (see @fig-sem and table S4). While we found that ROS production increased with metabolic capacity (β = `r format_dec(mean(coeff_potential_ros), 3)`, 95% CI = [`r format_dec(quantile(coeff_potential_ros, 0.025), 3)`, `r format_dec(quantile(coeff_potential_ros, 0.975), 3)`], _pMCMC_ `r format_p(pmcmc(coeff_potential_ros), 3, equal = TRUE)`), ROS was not related to  oxidative damage (see @fig-sem and table S4).

@@

To ensure that neurons were not unintentionally lost during homogenization, we performed a pilot study where we euthanized four lizards and prepared medial cortex homogenates using the same procedures as before. However, to ensure we could identify neurons, homogenates here were dyed with marker that specifically targeted neuron nuclei (Farrow et al. 2021; Storks et al. 2023). We employed fluorescence microscopy and flow cytometry to check for the presence of neurons in the homogenates.

Euthanasia and homogenization for each of brain region followed the procedures outlined above (see _Methods 2: Flow cytometry_). Before staining the samples, cells were also fixed and permeabilized as described previously (see _Methods 2: Flow cytometry_). Following permeabilization, we centrifuged the samples at 1000 RCF for 10 minutes and resuspended the pellet in 100 µL of a 1:100 dilution of NeuN + Alexa488 fluorescent conjugate to dye the neuronal nuclei (Farrow et al. 2021; Storks et al. 2023). The samples were incubated at +4 ºC overnight.

@@

Farrow LF, Andronicos NM, McDonald PG, Hamlin AS. 2021. Quantitative determination of neuronal size and density using flow cytometry. _Journal of Neuroscience Methods_, **352**, 109081.

Storks L, Powell BJ, Leal M. 2023. Peeking inside the lizard brain: neuron numbers in Anolis and its implications for cognitive performance and vertebrate brain evolution. _Integrative and Comparative Biology_, **63(1)**, 223-237.

>>>>

---
title: "Prenatal conditions do not affect brain physiology and learning in a lizard"
docx: 
bibliography: "./bib/refs.bib"
csl: "./bib/the-journal-of-experimental-biology.csl"
reference-doc: "./bib/tmpl.docx"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig.    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

## Keywords
cognition, corticosterone, learning, mitochondria, reptiles, incubation temperature

@@

We fitted a multivariate structural equation model (SEM) using _brms_ to explore direct and indirect links between mitochondrial function and learning. The model was structured based on theoretical expectations shown in @fig-sem. Any missing data was imputed during model fitting using data augmentation [@noble2021planned], but this was largely restricted to DNA damage and lipid peroxidation. To obtain a measure of learning performance we extracted the posterior distribution of individual learning slopes (i.e., changes in error over time) by including a random trial slope for each lizard. Learning slopes and the rest of variables, were standardized as before (see above). We assumed a Gaussian error distribution for all the variables in this model. Factors found to be non-significant in the univariate models were excluded. Clutch identity remained a random factor in our SEM model. Direct, indirect, and total effects were derived from posterior estimates [@kline2005principles] (see Table S4).

@@

We found that the lizards improved their choices over time, making less mistakes as the trials progressed (see @fig-learning and Table S1). However, the learning rate was not influenced by CORT, temperature, or their interaction (see contrasts in Table S2).

Mitochondrial density, metabolic capacity, ROS, DNA damage, and lipid peroxidation were also not affected by prenatal conditions (see @fig-results_mit and Table S2). We did not find effects of sex and age on mitochondrial physiology (see Table S3 and Tables S5-S9).

@@

Our SEM showed that mitochondrial physiology was unrelated to learning abilities (see @fig-sem and table S4). While we found that ROS production increased with metabolic capacity (β = `r format_dec(mean(coeff_potential_ros), 3)`, 95% CI = [`r format_dec(quantile(coeff_potential_ros, 0.025), 3)`, `r format_dec(quantile(coeff_potential_ros, 0.975), 3)`], _pMCMC_ `r format_p(pmcmc(coeff_potential_ros), 3, equal = TRUE)`), ROS was not related to  oxidative damage (see @fig-sem and Table S4).

@@

To ensure that neurons were not unintentionally lost during homogenization, we performed a pilot study where we euthanized four lizards and prepared medial cortex homogenates using the same procedures as before. However, to ensure we could identify neurons, homogenates here were dyed with marker that specifically targeted neuron nuclei (Farrow et al., 2021; Storks et al., 2023). We employed fluorescence microscopy and flow cytometry to check for the presence of neurons in the homogenates.

Euthanasia and homogenization for each of brain region followed the procedures outlined above (see _Methods 2: Flow cytometry_). Before staining the samples, cells were also fixed and permeabilized as described previously (see _Methods 2: Flow cytometry_). Following permeabilization, we centrifuged the samples at 1000 RCF for 10 minutes and resuspended the pellet in 100 µL of a 1:100 dilution of NeuN + Alexa488 fluorescent conjugate to dye the neuronal nuclei (Farrow et al., 2021; Storks et al., 2023). The samples were incubated at +4 ºC overnight.

@@

**Farrow, L. F., Andronicos, N. M., McDonald, P. G. and Hamlin, A. S.** (2021). Quantitative determination of neuronal size and density using flow cytometry. _Journal of Neuroscience Methods_, **352**, 109081.

**Storks, L., Powell, B. J. and Leal, M. (2023)**. Peeking inside the lizard brain: neuron numbers in Anolis and its implications for cognitive performance and vertebrate brain evolution. _Integrative and Comparative Biology_, **63(1)**, 223-237.
